---
title: "hw4"
output: html_document
author: "Matej Popovski"
date: "`r Sys.Date()`"
---

```{r, echo = FALSE}
knitr::opts_chunk$set(warnings = FALSE, message = FALSE)
```

> Geospatial Datasets 

a. NYC Building Footprints
Type: Vector
Geometry: Polygon
Reasoning: Building footprints represent the physical outline of buildings, which are best represented as polygons in vector data.

b. Africa Population 2020
Type: Raster
Reasoning: Population data over a large area like a continent is typically distributed across a grid, with each pixel representing population density or count. This is raster-based.

c. Himalayan Glacial Lakes
Type: Vector
Geometry: Polygon or Point
Reasoning: Each lake can be represented as a polygon (to show the surface area) or sometimes as a point (if only the location is needed). Most detailed datasets use polygons.

d. Wisconsin EV Charging
Type: Vector
Geometry: Point
Reasoning: Charging stations are individual locations, best represented as points in a vector format.

e. Zion Elevation
Type: Raster
Reasoning: Elevation is continuous data and is best represented as a raster (like a Digital Elevation Model), where each cell has an elevation value.




```{r}




```













































