---
title: "Problem Set 4 Coding Solutions"
subtitle: "STAT 436, Spring 2025"
output: rmdformats::readthedown
---

```{r setup, include=FALSE}
library(knitr)
opts_chunk$set(echo = TRUE, message = FALSE, warning = FALSE, cache = T, fig.align = "center")
```

```{r, echo = FALSE}
    library(tidyverse)
    library(sf)
    my_theme <- theme_classic() +
      theme(
        panel.background = element_rect(fill = "transparent", colour = NA),
        panel.grid.minor = element_blank(),
        axis.ticks = element_blank(),
        plot.background = element_rect(fill = "transparent", colour = NA)
      )
    theme_set(my_theme)
```


# Geospatial Datasets

For each of the datasets below, specify whether it is in a vector or raster data
format. If it is in a vector data format, explain which types of geometries it
contains (e.g., a point or linestring). Explain your reasoning.

a. [NYC Building Footprints](https://uwmadison.box.com/shared/static/qfmrp9srsoq0a7oj0e7xmgu5spojr33e.geojson)

Note: We have included visualizations, but these were not required for credit.

Vector. Each building is a multipolygon geometry, see the `Geometry type` field.
Each row gives the coordinates of the boundaries of the building.

```{r}
buildings <- read_sf("https://uwmadison.box.com/shared/static/qfmrp9srsoq0a7oj0e7xmgu5spojr33e.geojson")
buildings
```

```{r}
library(tmap)
tm_shape(buildings) +
    tm_polygons()
```

b. [Africa Population 2020](https://github.com/krisrs1128/stat436_s24/raw/main/data/afripop2020.tif)

Raster. This is shown in the `RasterBrick` data class below. Population is given
for each 0.1666 degrees of the longitude/latitude on the continent.

```{r}
library(raster)
afripop <- brick("https://github.com/krisrs1128/stat436_f24/raw/refs/heads/main/data/afripop2020.tif")
afripop
```

```{r}
tm_shape(log(1 + afripop)) +
    tm_raster(palette = c('#ffffd9','#edf8b1','#c7e9b4','#7fcdbb','#41b6c4','#1d91c0','#225ea8','#0c2c84'))
```

c. [Himalayan Glacial Lakes](https://raw.githubusercontent.com/krisrs1128/stat436_s23/main/data/GL_3basins_2015.topojson)

Vector. Each lake is a polygon.

```{r}
read_sf("https://raw.githubusercontent.com/krisrs1128/stat436_s23/main/data/GL_3basins_2015.topojson") |>
    mutate(GL_ID = fct_reorder(GL_ID, Area)) |>
    slice_max(Area, n = 20) |>
    tm_shape() +
    tm_polygons(col = "#72B6F2") +
    tm_facets("GL_ID", ncol = 5) +
    tm_scale_bar()
```

d. [Wisconsin EV Charging ](https://github.com/krisrs1128/stat436_s24/raw/main/data/ev.geojson)

Vector. Each station is a point geometry, and the rows give their latitude and
longitude coordinates. Can you recognize the outlines of Wisconsin from this?
Which regions have the most stations?.

```{r}
ev <- read_sf("https://github.com/krisrs1128/stat436_s24/raw/main/data/ev.geojson")
ev
```

```{R}
tm_shape(ev) +
    tm_dots()
```

e. [Zion Elevation](https://github.com/krisrs1128/stat436_s24/raw/main/data/landsat.tif)

Raster. This is because the object is class `RasterLayer`. We have elevation at
30 meter resolution across the park.

```{r}
zion <- brick("https://github.com/krisrs1128/stat436_s24/raw/main/data/landsat.tif") |>
    subset(4)
zion
```

```{r}
tm_shape(zion) +
    tm_raster(palette = c('#fcfbfd','#efedf5','#dadaeb','#bcbddc','#9e9ac8','#807dba','#6a51a3','#4a1486'))
```

# CalFresh Comparison

```{r, unemployment_read}
library(tsibble)
library(feasts)
unemployment <- read_sf("https://raw.githubusercontent.com/krisrs1128/stat436_s25/refs/heads/main/data/unemployment.geojson") |>
  mutate(date = yearmonth(date))
focus <- c("Colusa", "Monterey", "San Francisco", "Imperial", "Los Angeles", "San Joaquin")
head(unemployment, 3)
```
    
c. Provide code to answer: For each county and year, what was the maximum unemployment rate across all months?

```{r}
unemployment |>
  group_by(county, year) |>
  summarise(max_unemployed = max(unemployment, na.rm = TRUE))
```
    
d. Provide code to implement one of the figures above. You may assume the result from part (c) and do not need to match color or text themes.

```{r unemployment_figures, echo = TRUE, fig.width = 8, fig.height = 5, out.width = "0.7\\textwidth", fig.align = "center"}
library(scico)
library(tmap) 
    
p1 <- unemployment |>
  as_tsibble(county, date) |>
  filter(county %in% focus) |>
  gg_season(unemployment, linewidth = 1) +
  guides(color = guide_legend(override.aes = list(linewidth = 2))) +
  facet_wrap(~ reorder(county, unemployment), ncol = 2) +
  labs(title = "Approach 1") +
  theme(
    title = element_text(size = 18),
    axis.title = element_text(size = 16),
    axis.text.y = element_text(size = 14),
    axis.text.x = element_text(size = 10),
    strip.text = element_text(size = 16),
    legend.text = element_text(size = 14),
    legend.position = "right"
  )

p2 <- ggplot(unemployment) +
  geom_tile(
    aes(
      as_date(date),
      reorder(county, unemployment, na.rm = TRUE),
      fill = sqrt(unemployment),
      col = sqrt(unemployment)
    ),
    linewidth = 1
  ) +
  scale_x_date(expand = c(0, 0)) +
  scale_fill_scico(palette = "roma") +
  scale_color_scico(palette = "roma") +
  labs(x = "Date", y = "County", fill = "sqrt(Unemployment %)", color = "sqrt(Unemployment %)", title = "Approach 2") +
  theme(
    title = element_text(size = 18),
    axis.title = element_text(size = 18),
    axis.text.y = element_text(size = 8),
    axis.text.x = element_text(size = 16),
    legend.text = element_text(size = 16),
    legend.title = element_text(size = 18)
  )

p3 <- unemployment |>
  group_by(county, year) |>
  summarise(max_unemployed = max(unemployment, na.rm = TRUE)) |>
  tm_shape() +
  tm_polygons(col = "max_unemployed", title = "Max. Unemployed") +
  tm_facets("year", ncol = 4) +
  tm_layout(
    main.title = "Approach 3",
    title.size = 14
  )

p1
```
    
```{r, echo = TRUE, fig.width = 10, fig.height = 6.25, out.width = "0.7\\textwidth", fig.align = "center", dpi = 300}
p2
```
    
```{r, echo = TRUE, fig.width = 6, fig.height = 3, out.width = "0.7\\textwidth", fig.align="center"}
p3
```


# HIV Network

a. Provide code to generate a node-link visualization similar to that in Figure 14. Make at least one style customization that improves the visualization. Justify your choice.

```{r, echo = TRUE, warning = FALSE, message = FALSE}
    library(tidygraph)
    nodes <- read_csv("https://github.com/krisrs1128/stat436_s25/raw/refs/heads/main/data/hiv_nodes.csv")
    edges <- read_csv("https://github.com/krisrs1128/stat436_s25/raw/refs/heads/main/data/hiv_edges.csv")
    G <- tbl_graph(nodes, edges)

    library(ggraph)
    ggraph(G) +
      geom_edge_link(colour = "#d3d3d3", alpha = 0.9, width = 2) +
      geom_node_text(aes(label = name)) +
      theme_void()
```
    
b. The figure below shows the node-link visualization for a simple graph. Sketch its exact adjacency
matrix.

```{r}
tibble(
  node = 1:5,
  `1` = c(0, 1, 1, 0, 1),
  `2` = c(1, 0, 0, 1, 0),
  `3` = c(1, 0, 0, 0, 0),
  `4` = c(0, 1, 0, 0, 0),
  `5` = c(1, 0, 0, 0, 0)
) %>% 
  column_to_rownames(var = "node")
```


c. The figure below shows the node-link visualization for a more complex graph. How do you expect its
corresponding adjacency matrix visualization to look? Provide a rough sketch and a one-sentence
justification.

The three cluster structure of the node-link visualization would result in an adjacency matrix with three distinct, dense blocks of 1's, with a smattering of 1's not in those blocks representing the small number of connections across clusters.
    
# Beijing Air Pollution

This problem studies daily air pollution and weather
patterns in Beijing from 2010 to 2014. Each row in this dataset gives
measurements for three variables, `pollution`, `temp`, and `wndspd` over time
for a single day, indexed by `date`. For example the `temp_13` column gives the
temperature at 1pm. Note that both `pollution` and `wndspd` have been
$\log\left(1 + x\right)$ transformed.

    a. Cluster all the days using a hierarchical clustering. Be sure to scale
    the columns in advance, so that no single variable has undue influence on
    the result. Cut the tree so that we are left with 10 clusters, and provide a
    ribbon plot to visualize several quantiles associated with each cluster's
    typical pollution, temperature, and windspeed trajectories. visualization to
    understand the centroids of each cluster. Briefly interpret the main
    patterns you observe.
    
```{r}
pollution <- read_csv("https://raw.githubusercontent.com/krisrs1128/stat436_s25/refs/heads/main/data/pollution_wide.csv")

fit <- pollution %>%
  select(-date) %>%
  scale() %>%
  dist() %>%
  hclust()
```
 
```{r}
ribbon_data <- pollution %>%
  mutate(cluster = cutree(fit, k = 10)) %>%
  pivot_longer(contains("_")) %>%
  separate(name, c("variable", "hour"), convert = TRUE) %>%
  group_by(cluster, variable, hour) %>%
  summarise(
    q1 = quantile(value, 0.25),
    median = median(value),
    q3 = quantile(value, 0.75)
  )

cluster_order <- ribbon_data %>%
  filter(variable == "pollution") %>%
  group_by(cluster) %>%
  summarise(m = mean(median)) %>%
  arrange(-m) %>%
  pull(cluster)

ribbon_data <- ribbon_data %>%
  mutate(cluster = factor(cluster, levels = cluster_order))

ggplot(ribbon_data) +
  geom_ribbon(aes(hour, ymin = q1, ymax = q3)) +
  geom_line(aes(hour, median)) +
  facet_grid(variable ~ cluster, scale = "free_y")
```