---
title: "HW 1 "
output: html_notebook
London Olympics
 author:
  - name: Matej Popovski
---


```{r}
knitr::opts_chunk$set(cache = FALSE, message = FALSE, warning = FALSE, echo = TRUE)
```

```{r}
# Load necessary libraries
library(dplyr)
library(ggplot2)
#library(RColorBrewer)
#display.brewer.all()
```

a.	 Create a layered display that shows
 (i) the ages of athletes across sports and 
 (ii) the average age within each sport.
 
```{r}

# Load the olympics package
olympics <- read_csv("https://uwmadison.box.com/shared/static/rzw8h2x6dp5693gdbpgxaf2koqijo12l.csv", show_col_types = FALSE)
head(olympics)
```
 b. Sort the sports from lowest to highest average age.
 
```{r}
# Project the original table to columns Name, Age, Sport
olympics_sportists <- olympics|> select(Name,Age,Sport)
# Display the first few rows of the projected table
head(olympics_sportists)
```




```{r}
# Create table containing average age for all sports
avg_age_by_sport <- olympics_sportists |>
  group_by(Sport) |>
  summarise(Avg_age = mean(Age, na.rm = TRUE)) |>
  arrange(Avg_age)
head(avg_age_by_sport)
```


```{r, fig.height = 6.7, fig.width = 4.5}

# Create boxplot for age in various sports in LOndon Oimpics

ggplot(olympics_sportists, aes(x = Sport, y = Age)) +
  geom_boxplot() +
  labs(title = "Age Distribution of Athletes by Sport",
       x = "Sport",
       y = "Age") +
  theme(axis.text.x=element_text(angle=90), panel.grid.major.x=element_blank())

ggplot(olympics_sportists, aes(x = Sport, y = Age)) +
  geom_point() +
  labs(title = "Age Distribution of Athletes by Sport",
      x = "Sport", y = "Age") +
  theme(axis.text.x=element_text(angle=90), panel.grid.major.x=element_blank())
```

```{r, fig.height = 6.7, fig.width = 4.5}
# Create boxplot for age in various sports in LOndon Oimpics

ggplot(olympics_sportists, aes(reorder(Sport,-Age, median), y = Age)) +
  geom_boxplot() +
  labs(title = "Age Distribution of Athletes by Sport",
       x = "Sport",
       y = "Age") +
  theme(axis.text.x=element_text(angle=90), panel.grid.major.x=element_blank())

ggplot(olympics_sportists, aes(reorder(Sport,-Age, mean), y = Age)) +
  geom_point() +
  labs(title = "Age Distribution of Athletes by Sport",
      x = "Sport", y = "Age") +
  theme(axis.text.x=element_text(angle=90), panel.grid.major.x=element_blank())
```



c.  Develop one new question based on these data. What makes you interested in it?
 Provide a visualization that supports the comparisons needed to arrive at an answer.

```{r  }
# Create the scatter plot with a trend line
# ggplot(avg_age_by_sport, aes(x = Avg_age, y = Sport)) +
  ggplot(avg_age_by_sport, aes(x = reorder(Sport,-Avg_age, median), y=Avg_age)) +
  geom_point() +
  geom_smooth(method = "lm", se = FALSE) +
  labs(title = "Average Age of Athletes by Sport (London 2012 Olympics)",
       x = "Average Age",
       y = "Sport") +
  theme(axis.text.x=element_text(angle=90), panel.grid.major.x=element_blank())
```
 
 
 kinezot
 
```{r}
# Create a bar plot
ggplot(avg_age_by_sport, aes(x = reorder(Sport, -Avg_age), y = Avg_age)) +
  geom_bar(stat = "identity", fill = "steelblue") +
  theme(axis.text.x = element_text(angle = 90, hjust = 1)) +
  labs(
    x = "Sport",
    y = "Average Age",
    title = "Average Age of Athletes by Sport in London 2012 Olympics"
  ) 

```
 
 kinez   da se dopravi so sitnici okolu bars  od prethodno
 
```{r}
# Create a bar plot with swapped axes
ggplot(avg_age_by_sport, aes(x = Avg_age, y = reorder(Sport, Avg_age))) +
  geom_bar(stat = "identity", fill = "steelblue") +
  labs(
    x = "Average Age",
    y = "Sport",
    title = "Average Age of Athletes by Sport in London 2012 Olympics"
  ) +
  theme(axis.text.y = element_text(hjust = 1))  # Adjust y-axis text alignment
```
 podobruvananja
 
```{r}
# Example: Set x-axis limits for the bar plot
ggplot(avg_age_by_sport, aes(x = Avg_age, y = reorder(Sport, Avg_age))) +
  geom_bar(stat = "identity", fill = "steelblue") +
  labs(
    x = "Average Age",
    y = "Sport",
    title = "Average Age of Athletes by Sport in London 2012 Olympics"
  ) +
  xlim(0, 40) +  # Set x-axis limits
  theme_minimal()
```
```{r}
# Example: Add breaks and format labels for the x-axis
ggplot(avg_age_by_sport, aes(x = Avg_age, y = reorder(Sport, Avg_age))) +
  geom_bar(stat = "identity", fill = "steelblue") +
  labs(
    x = "Average Age",
    y = "Sport",
    title = "Average Age of Athletes by Sport in London 2012 Olympics"
  ) +
  scale_x_continuous(breaks = seq(0, 40, by = 5)) +  # Add breaks every 5 units
  theme_minimal()
```
 
 kinez  broj na outliers per sport  kolku ima postari ili pomladi od tie vo IQR
 
 
```{r}

# Function to identify outliers
find_outliers <- function(x) {
  Q1 <- quantile(x, 0.25, na.rm = TRUE)
  Q3 <- quantile(x, 0.75, na.rm = TRUE)
  IQR <- Q3 - Q1
  lower_bound <- Q1 - 1.5 * IQR
  upper_bound <- Q3 + 1.5 * IQR
  outliers <- x[x < lower_bound | x > upper_bound]
  return(outliers)
}

# Calculate the number of outliers per sport
outliers_by_sport <- olympics_sportists %>%
  group_by(Sport) %>%
  summarise(
    Number_of_Outliers = length(find_outliers(Age))
  ) %>%
  arrange(desc(Number_of_Outliers))  # Sort by number of outliers

# Print the table
print(outliers_by_sport)
```
 
 
 
```{r, fig.height = 6.7, fig.width = 6 }
# Create a bar plot for outliers
ggplot(outliers_by_sport, aes(x = reorder(Sport, -Number_of_Outliers), y = Number_of_Outliers)) +
  geom_bar(stat = "identity", fill = "tomato") +
  # theme(axis.text.x = element_text(angle = 90, hjust = 1)) +
  labs(
    x = "Sport",
    y = "Number of Outliers",
    title = "Distribution of Outliers by Sport in London 2012 Olympics"
  ) +
  theme(axis.text.x = element_text(angle = 90, hjust = 1))
```
 Outliers can indicate unusual age distributions in certain sports, which might be worth further investigation.
 
 
 
 
 
 
 