---
title: "Topic Modeling Hotel Reviews with LDA"
output: html_document
date: "`r Sys.Date()`"
---

## Summary

This R Markdown document demonstrates the use of Latent Dirichlet Allocation (LDA) to extract underlying topics from a set of hotel reviews. The implementation involves loading and preprocessing text data, transforming it into a document-term matrix, and then fitting an LDA model using the `topicmodels` package. The results are visualized to explore common terms in each topic and investigate how topics vary across positive and negative review ratings.

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```
# Sets global chunk options for R Markdown

```{r, echo = FALSE, message = FALSE, warning = FALSE}
library(knitr)
library(tidyverse)
library(topicmodels)
library(tidytext)

# Sets a minimalist theme for plots
th <- theme_minimal() + 
  theme(
    panel.grid.minor = element_blank(),
    panel.background = element_rect(fill = "#f7f7f7"),
    panel.border = element_rect(fill = NA, color = "#0c0c0c", size = 0.6),
    axis.text = element_text(size = 14),
    axis.title = element_text(size = 16),
    legend.position = "bottom"
  )
theme_set(th)
```
# Loads required libraries and sets a custom plotting theme

```{r}
# Load hotel reviews dataset
reviews <- read_csv("https://github.com/krisrs1128/stat436_f24/raw/refs/heads/main/data/reviews.csv")

# Tokenize text into words and remove stopwords
reviews_dtm <- reviews |>
  unnest_tokens(word, Review) |>                      # Converts reviews into individual lowercase words
  filter(!(word %in% stop_words$word)) |>             # Removes common stopwords
  count(document, word) |>                            # Counts word frequencies per review
  cast_dtm(document, word, n)                         # Converts to a DocumentTermMatrix for LDA
```

```{r}
# Fit LDA model with 8 topics and a fixed seed for reproducibility
fit <- LDA(reviews_dtm, k = 8, control = list(seed = 123))
```

```{r}
# Extract beta matrix (topic-term probabilities)
topics <- tidy(fit, matrix = "beta")
```

```{r}
# Get top 10 terms per topic based on beta value
top_terms <- topics |>
  group_by(topic) |>
  slice_max(beta, n = 10) |>                          # Selects top 10 words per topic
  ungroup() |>
  arrange(topic)
```

```{r}
# Plot the top terms per topic
top_terms |>
  mutate(term = reorder_within(term, beta, topic)) |>
  ggplot(aes(beta, term, fill = factor(topic))) +
  geom_col(show.legend = FALSE) +
  facet_wrap(~ topic, scales = "free") +              # One plot for each topic
  scale_y_reordered()
```

```{r}
# Extract gamma matrix (document-topic probabilities)
reviews_topics <- tidy(fit, matrix = "gamma") |> 
  # Ensure the `document` column in `reviews` is a character to match the type in gamma matrix
  left_join(reviews |> mutate(document = as.character(document)), by = "document")

```

```{r}
# Plot average topic probabilities by review rating
reviews_topics |>
  group_by(topic, Rating) |>
  summarize(avg_gamma = mean(gamma)) |>               # Computes average topic strength per rating
  ggplot(aes(Rating, avg_gamma, fill = factor(topic))) +
  geom_col(position = "dodge") +
  labs(y = "Avg. Topic Probability", x = "Rating")
```

