## Summary of Implementation Strategy

This analysis demonstrates the application of UMAP to a single-cell genomics dataset containing PBMC (Peripheral Blood Mononuclear Cells) data. The workflow includes loading and preprocessing the dataset, applying dimensionality reduction using the UMAP method from the `embed` package, and visualizing the 2D embedding. The strategy emphasizes preserving local structure in the data, which is essential for identifying potential biological subpopulations. The use of `tidymodels` helps structure the preprocessing and modeling pipeline in a reproducible manner.


```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```


# ---- Section 1: Description of the next code block ----
---
title: "Untitled"
output: html_document
date: "`r Sys.Date()`"
---

```{r, echo = FALSE}
knitr::opts_chunk$set(warnings = FALSE, message = FALSE)
```

```{r}
library(tidyverse)
library(embed)

# ---- Section 2: Description of the next code block ----
library(tidymodels)
th <- theme_minimal() +
    theme(
        panel.grid.minor = element_blank(),
        panel.background = element_rect(fill = "#f7f7f7"),
        panel.border = element_rect(fill = NA, color = "#0c0c0c"),
        axis.text = element_text(size = 14),
        axis.title = element_text(size = 16),
        strip.text = element_text(size = 16),
        legend.position = "bottom"
    )
theme_set(th)
```

# ---- Section 3: Description of the next code block ----

[Single-Cell Genomics]  In this problem, we will apply UMAP to a
[dataset](https://uwmadison.box.com/shared/static/ai539s30rjsw5ke4vxbjrxjaiihq7edk.csv)
of Peripheral Blood Mononuclear Cells (PBMC) released by 10X Genomics. The first
column, `cell_tag`, gives an identifier for each cell in the dataset. All other
columns are molecules that were detected in that cell. For example, CD74 is a
molecule often found on the surface of T-cells.
    
a. Define a tidymodels `recipe` that specifies that UMAP should be performed
with the parameters `learn_rate = 0.1` and `neighbors = 5`. There is no need to
normalize these data, as they have been normalized in advance using methods
tailored to single-cell genomics data.


# ---- Section 4: Description of the next code block ----
```{r}
pbmc <- read_csv("https://uwmadison.box.com/shared/static/ai539s30rjsw5ke4vxbjrxjaiihq7edk.csv")
# umap_rec <-
```

b. Compute the UMAP embeddings across cells. Color points in by their value of
the GNLY molecule.

