Loading Libraries: tidyverse for data manipulation and ggplot2 for visualization.

```{r}
# Load necessary libraries
library(tidyverse)
library(ggplot2)
```

Reading Data: The dataset is read from the provided URL using read_csv

```{r}
# Read the dataset
ikea <- read_csv("https://uwmadison.box.com/shared/static/iat31h1wjg7abhd2889cput7k264bdzd.csv", show_col_types = FALSE)
print(ikea)
```
Plot Creation: The ggplot function initializes the plot. The aes function maps category to the x-axis, price to the y-axis, and online_purchase given by sellable_online to the color of the points.


geom_jitter: Jitters the points to reduce overlap. The width parameter controls the amount of horizontal jitter.

scale_y_log10: Applies a log scale to the y-axis for prices.

Adds labels and a title to the plot.

Applies a clean and minimal theme to the plot.

```{r}
# Create a plot with jittered points and log-scaled price
ggplot(ikea, aes(x = category, y = price, color = sellable_online)) +
  geom_jitter(width = 0.2, height = 0, size = 2) +
  scale_y_log10() +
  labs(title = "IKEA Furniture Prices by Category",
       x = "Furniture Category",
       y = "Price (Log Scale)",
       color = "Sellable Online") +
  theme_minimal()
```


```{r}
library(gapminder)
data(gapminder)
data
```

```{r}

```
















