---
title: "hw5"
author: "Matej Popovski"
date: "`r Sys.Date()`"
output: html_document
---
  
```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, message = FALSE, warning = FALSE)
library(tidyverse)
library(janitor)
library(recipes)
library(ggrepel)
library(patchwork)
```

## Data Loading and Cleaning

```{r load-data}
# Load and clean the dataset
air_quality <- read_csv("Air_Quality.csv") %>%
  clean_names() %>%
  filter(!is.na(data_value))
air_quality
# Focus on key variables
cleaned_data <- air_quality %>%
  select(geo_place_name, name, data_value) %>%
  pivot_wider(names_from = name, values_from = data_value, values_fn = mean) %>%
  drop_na()
cleaned_data
# Save neighborhood labels for plotting
location_labels <- cleaned_data$geo_place_name
location_labels

# Numeric data for PCA
numeric_data <- cleaned_data %>%
  select(-geo_place_name)
numeric_data
```

## PCA Preprocessing

```{r pca-prep}
rec <- recipe(~ ., data = numeric_data) %>%
  step_normalize(all_numeric_predictors()) %>%
  step_pca(all_numeric_predictors(), num_comp = 2)
pca_data <- prep(rec) %>% bake(new_data = NULL)
pca_df <- bind_cols(pca_data, geo_place_name = location_labels)
pca_df
```

## PCA Visualization 1: Neighborhoods by Air Quality

```{r pca-plot}
ggplot(pca_df, aes(x = PC1, y = PC2, label = geo_place_name)) +
  geom_point(color = "steelblue", alpha = 0.7) +
  geom_text_repel(size = 2.5) +
  labs(
    title = "PCA of NYC Neighborhoods Based on Air Quality",
    subtitle = "Pollution profiles reduced via PCA",
    x = "Principal Component 1",
    y = "Principal Component 2"
  ) +
  theme_minimal()
```
=====
copilot MM

```{r}
# Standardize the data
scaled_data <- scale(numeric_data)
#scaled_data
```





```{r}


# Perform PCA
pca_full <- prcomp(numeric_data, center = TRUE, scale. = TRUE)

# Summary of PCA to see the explained variance
summary(pca_full)

# Extract PCA loadings
loadings <- pca_full$rotation
print("PCA Loadings:")
print(loadings)
#install.packages("ggbiplot")
library(ggbiplot)
ggbiplot(pca_full, labels = rownames(numeric_data), ellipse = TRUE, circle = TRUE)

# Visualize the PCA
biplot(pca_full, scale = 0.9)

```

popravka so iminja

```{r}
# Perform PCA
pca_full <- prcomp(numeric_data, center = TRUE, scale. = TRUE)

# Summary of PCA to see the explained variance
summary(pca_full)

# Extract PCA loadings
loadings <- pca_full$rotation
print("PCA Loadings:")
print(loadings)
#install.packages("ggbiplot")
library(ggbiplot)
ggbiplot(pca_full, labels = pca_df$geo_place_name, ellipse = TRUE, circle = TRUE)
#ggbiplot(pca_full, labels = rownames(numeric_data), ellipse = TRUE, circle = TRUE)

```
=== ===
improve plot I

```{r}
library(ggbiplot)

# Enhanced ggbiplot with better visibility
ggbiplot(pca_full, 
         labels = pca_df$geo_place_name,  # Add region names
         ellipse = TRUE, 
         circle = TRUE, 
         varname.size = 2,  # Increase arrow label font size
         varname.adjust = 1.5,  # Adjust arrow label positions slightly
         var.scale = 1) +  # Scale down arrow length
  theme_minimal() +  # Cleaner background for better readability
  theme(text = element_text(size = 10),  # Increase text size for labels
        legend.position = "right")  # Move legend for better alignment
```
====
 


improve plot II

```{r}
# Assign cluster numbers (based on k-means or another clustering method)
clusters <- kmeans(pca_full$x[, 1:2], centers = 3)$cluster
ggbiplot(pca_full, groups = as.factor(clusters), ellipse = TRUE, circle = TRUE) +
  scale_color_manual(values = c("red", "blue", "green"))  # Customize cluster colors
```
===

reduce # variables

```{r}
# Identify the most influential variables
important_variables <- rownames(loadings)[apply(abs(loadings[, 1:2]), 1, max) > 0.5]  # Threshold of 0.5

# Filter PCA loadings for selected variables
ggbiplot(pca_full, labels = pca_df$geo_place_name, ellipse = TRUE, circle = TRUE, 
         varname.limit = length(important_variables))
```

====

K-means

```{r}
# Perform K-Means Clustering
set.seed(123)
clusters <- kmeans(pca_full$x[, 1:2], centers = 3)  # Use PC1 and PC2, 3 clusters

# Add cluster labels to the PCA data
scores <- data.frame(PC1 = pca_full$x[, 1], PC2 = pca_full$x[, 2], Cluster = as.factor(clusters$cluster))

# Visualize Clusters
library(ggplot2)
ggplot(scores, aes(x = PC1, y = PC2, color = Cluster)) +
  geom_point(size = 3) +
  labs(title = "Clusters in PC1-PC2 Space", x = "Principal Component 1", y = "Principal Component 2") +
  theme_minimal()
```
=== ===
```{r}
# Create a data frame for PCA scores
scores <- data.frame(PC1 = pca_full$x[, 1], 
                     PC2 = pca_full$x[, 2], 
                     Region = pca_df$geo_place_name)  # Add regional names

# Plot the PCA scatterplot with regional names
library(ggplot2)
ggplot(scores, aes(x = PC1, y = PC2, label = Region)) +
  geom_point(color = 'blue', size = 3) +  # Points for regions
  geom_text(aes(label = Region), hjust = 1.1, vjust = -0.5, size = 3, color = "darkred") +  # Regional names
  labs(title = "PCA Scatterplot with Regional Names", 
       x = "Principal Component 1 (PC1)", 
       y = "Principal Component 2 (PC2)") +
  theme_minimal()
```
==
```{r}
# Perform K-Means Clustering
set.seed(123)  # Ensures reproducibility
clusters <- kmeans(pca_full$x[, 1:2], centers = 3)  # Use PC1 and PC2, with 3 clusters

# Create a data frame for PCA scores, clusters, and regional names
scores <- data.frame(PC1 = pca_full$x[, 1], 
                     PC2 = pca_full$x[, 2], 
                     Region = pca_df$geo_place_name,  # Regional names
                     Cluster = as.factor(clusters$cluster))  # Cluster labels

# Plot the PCA scatterplot with regional names and color by clusters
library(ggplot2)
ggplot(scores, aes(x = PC1, y = PC2, label = Region, color = Cluster)) +
  geom_point(size = 3) +  # Points for regions, colored by cluster
  geom_text(aes(label = Region), hjust = 1.1, vjust = -0.5, size = 3, color = "black") +  # Regional names
  labs(title = "PCA Scatterplot with Clusters and Regional Names", 
       x = "Principal Component 1 (PC1)", 
       y = "Principal Component 2 (PC2)", 
       color = "Cluster") +  # Cluster legend
  theme_minimal()
```
podobruvanje vo vidlivost
===
```{r}
# Perform K-Means Clustering
set.seed(123)  # Ensures reproducibility
clusters <- kmeans(pca_full$x[, 1:2], centers = 3)  # Use PC1 and PC2, with 3 clusters

# Create a data frame for PCA scores, clusters, and regional names
scores <- data.frame(PC1 = pca_full$x[, 1], 
                     PC2 = pca_full$x[, 2], 
                     Region = pca_df$geo_place_name,  # Regional names
                     Cluster = as.factor(clusters$cluster))  # Cluster labels

# Plot the PCA scatterplot with regional names and color by clusters
# Refined Plot for Better Label and Point Visibility
library(ggplot2)
library(ggrepel)  # For better label positioning

ggplot(scores, aes(x = PC1, y = PC2, color = Cluster)) +
  geom_point(size = 4, alpha = 0.8) +  # Larger points with some transparency
  geom_text_repel(aes(label = Region), 
                  size = 3.5,   # Adjust text size for readability
                  max.overlaps = 10,  # Limit overlaps with smart label positioning
                  box.padding = 0.3,  # Add space around labels
                  point.padding = 0.5,  # Add padding between points and labels
                  segment.color = "grey50",  # Add connecting lines
                  segment.size = 0.3) +  # Make lines thin and subtle
  labs(title = "PCA Scatterplot with Clusters and Regional Names", 
       x = "Principal Component 1 (PC1)", 
       y = "Principal Component 2 (PC2)", 
       color = "Cluster") +
  theme_minimal() +
  theme(legend.position = "right",  # Keep legend on the side
        plot.title = element_text(size = 14, face = "bold"),  # Improve title styling
        axis.title = element_text(size = 12))  # Improve axis label styling

```




===

Code Explanation
K-Means Clustering:
- Clusters regions based on their PC1 and PC2 scores.
- centers = 3: Creates 3 clusters (adjust this number based on your dataset).

Color Coding:
- The Cluster variable assigns distinct colors to each cluster, making it easy to visually group regions.

Text Labels:
- geom_text() displays the regional names next to their corresponding points.
- Use hjust and vjust to position the labels appropriately, avoiding overlap.

Cluster Legend:
- The color = Cluster mapping automatically creates a legend showing cluster colors.


Expected Outcome
- Clusters:
- Regions will be color-coded by cluster, highlighting groups with similar PCA profiles.
- For example, Port Richmond and Rockaway might fall within the same cluster.

Regional Names:
- Names will be displayed alongside points for easy identification.

Enhanced Visualization:
- The plot will clearly show clusters and regional names, improving interpretability.






=== ===

```{r}

ggbiplot(pca_full, labels = rownames(numeric_data), ellipse = TRUE, circle = TRUE)
```
```{r}
scores <- as.data.frame(pca_full$x)
scores$Neighborhood <- rownames(pca_full) # Add neighborhood names (optional)
scores
```


==== 
interaktiven plot

```{r}
# Load required libraries
#install.packages("plotly")
library(plotly)
library(ggplot2)


# Create a data frame for PCA scores (coordinates of neighborhoods)
scores <- as.data.frame(pca_full$x)
#scores$Neighborhood <- rownames(numeric_data) # Add neighborhood names (optional)
scores$Neighborhood <- cleaned_data$geo_place_name
# Create a data frame for loadings (variables contributing to PCA)
loadings <- as.data.frame(pca_full$rotation)
loadings$Variable <- rownames(loadings) # Add variable names

# Plot PCA scores (points) and loadings (arrows) interactively using plotly
p <- ggplot(scores, aes(x = PC1, y = PC2)) +
  geom_point(aes(text = paste("Neighborhood:",rownames(numeric_data) )), color = 'blue', size = 3) +
  geom_segment(data = loadings, aes(x = 0, y = 0, xend = PC1*10, yend = PC2*10), 
               arrow = arrow(length = unit(0.2, "cm")), color = 'red') +
  geom_text(data = loadings, aes(x = PC1*10, y = PC2*10, label = Variable), 
            hjust = 0.5, vjust = 0.5, color = 'darkred') +
  labs(title = "Interactive PCA Biplot",
       x = "Principal Component 1 (PC1)",
       y = "Principal Component 2 (PC2)") +
  theme_minimal()

# Convert ggplot to an interactive plotly object
interactive_plot <- ggplotly(p, tooltip = "text")

# Display the interactive plot
interactive_plot
```
=== ===

```{r}
# Updated ggplot with proper text mapping
p <- ggplot(scores, aes(x = PC1, y = PC2)) +
  geom_point(aes(text = Neighborhood), color = 'blue', size = 3) +  # Explicitly map text aesthetic
  geom_segment(data = loadings, aes(x = 0, y = 0, xend = PC1*10, yend = PC2*10), 
               arrow = arrow(length = unit(0.2, "cm")), color = 'red') +
  geom_text(data = loadings, aes(x = PC1*10, y = PC2*10, label = Variable), 
            hjust = 0.5, vjust = 0.5, color = 'darkred') +
  labs(title = "Interactive PCA Biplot",
       x = "Principal Component 1 (PC1)",
       y = "Principal Component 2 (PC2)") +
  theme_minimal()

# Convert ggplot to interactive plotly plot
interactive_plot <- ggplotly(p, tooltip = "text")  # Ensure tooltip uses the mapped `text` aesthetic

# Display the interactive plot
interactive_plot
```



==== ====

## Scree Plot: Variance Explained

### Initial Version

```{r scree-plot}
# Basic scree plot from PCA object
pca_full <- prcomp(numeric_data, scale. = TRUE)
pca_full

data.frame(PC = paste0("PC", 1:length(pca_full$sdev)),
           Variance = (pca_full$sdev^2) / sum(pca_full$sdev^2)) %>%
  ggplot(aes(x = PC, y = Variance)) +
  geom_col(fill = "darkgreen") +
  labs(title = "Scree Plot of Principal Components",
       x = "Principal Component", y = "Proportion of Variance Explained") +
  theme_minimal()
```


### Improved Version with Ordered PCs and Elbow Line

```{r scree-plot-improved}
# Scree plot with reordered PCs and horizontal elbow threshold
scree_df <- data.frame(
  PC = factor(paste0("PC", 1:length(pca_full$sdev)), levels = paste0("PC", 1:length(pca_full$sdev))),
  Variance = (pca_full$sdev^2) / sum(pca_full$sdev^2)
)

scree_df %>%
  ggplot(aes(x = PC, y = Variance)) +
  geom_col(fill = "forestgreen") +
  geom_hline(yintercept = 0.1, linetype = "dashed", color = "red") +
  labs(
    title = "Improved Scree Plot with Elbow Threshold",
    x = "Principal Component",
    y = "Proportion of Variance Explained"
  ) +
  theme_minimal()
```