---
title: "inclassdisc5-1"
output: html_document
date: "2025-02-18"
author: "Matej Popovski"
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

```{r}
# Load required libraries
library(tidyverse)  # For data manipulation and visualization
library(shiny)      # For building interactive web applications
library(DT)         # For rendering interactive tables in Shiny

# Set default ggplot theme
theme_set(theme_bw())

# Load the dataset and preprocess it
birds <- read_csv("https://raw.githubusercontent.com/krisrs1128/stat479_s22/main/exercises/data/birds.csv") |>
    separate(Species, c("genus", "species2"))  # Split the Species column into genus and species

# Summarize data by genus, computing mean and standard deviation for selected columns
bird_summaries <- birds |>
    group_by(genus) |>
    summarise(
        across(
            c("Asymmetry", "Ellipticity", "AvgLength (cm)", "Number of images", "Number of eggs"),
            list(
                MEAN = mean,  # Compute mean
                STANDARD_DEVIATION = sd  # Compute standard deviation
            )
        )
    ) |>
    arrange(-Ellipticity_MEAN)  # Sort in descending order of mean Ellipticity

# Function to create a scatterplot with brushed points highlighted
scatter <- function(df, selected) {
    palette <- c("#d4d4d4", "black")  # Define colors: grey for non-selected, black for selected points
    df |>
        mutate(selected_ = selected) |>  # Add a column indicating selection status
        ggplot() +
        geom_point(aes(Asymmetry, Ellipticity, color = as.factor(selected_))) +  # Scatterplot with color based on selection
        scale_color_manual(values = palette)  # Apply custom colors
}

# Define the UI (User Interface) components
ui <- fluidPage(
  titlePanel("Birds Brushed Scatterplot"),  # Application title
  plotOutput("plot", brush = "scatter_brush"),  # Scatterplot output with brushing enabled
  DTOutput("table")  # Data table output
)

# Define the server logic
server <- function(input, output) {
    selected <- reactiveVal(rep(TRUE, nrow(birds)))  # Initialize selection state for all points
     
    observeEvent(input$scatter_brush, {  # Listen for brushing events
        brushed_points <- brushedPoints(birds, input$scatter_brush, allRows = TRUE)  # Identify brushed points
        selected(brushed_points$selected_)  # Update selection state
    })

    output$plot <- renderPlot(scatter(birds, selected()))  # Render the updated scatterplot
    output$table <- renderDT(filter(birds, selected()))  # Display table of selected points
}

# Run the Shiny application
shinyApp(ui, server)
```

Summary:

This Shiny application visualizes bird egg characteristics by creating an interactive scatterplot of Asymmetry vs. Ellipticity. The app allows users to brush over a selection of points, highlighting them in a different color and displaying the corresponding species information in a table. The implementation follows a modular approach by defining a plotting function (scatter), setting up a ui (user interface) with a plot and a table, and implementing a server function to handle user interactions. The app uses reactiveVal to dynamically update the plot and table based on brushed selections.




